Fabric-Elasticity Relationships of Tibial Trabecular Bone are Similar in Osteogenesis Imperfecta and Healthy Individuals

### 2021-07-21 09_Article/Submission/Simon_2021_Bone_Draft.tex
Article corrections

\author[2,3,4]{Seyedmahdi Hosseinitabatabaei}

\author[2,3,4]{Bettina M. Willie}

@@

\address[3]{Department of Pediatric Surgery, McGill University, Montreal, Canada}

\address[4]{Department of Biomedical Engineering, McGill University, Montreal, Canada}

@@

	In the present study, the morphology of distal tibiae from 50 adults with OI were compared to 120 healthy controls using second generation HR-pQCT. Six cubic regions of interest (ROIs) were selected per individual in a common anatomical region. A first matching between OI and healthy control group was performed by selecting similar individuals to obtain identical mean and median age and sex distribution. It allowed us to perform a first morphometric analysis and compare the outcome with literature. Then, stiffness tensors of the ROIs were computed using  \si{\micro}FE and multiple linear regressions were performed with the Zysset-Curnier orthotropic fabric-elasticity model. %The linear regressions allowed to compare the two groups using five parameters.

@@

	Compared to healthy controls, we found the OI samples to have significantly lower BV/TV and trabecular number (Tb.N.), significantly higher trabecular separation (Tb.Sp.) and trabecular separation standard deviation (Tb.Sp.SD), but no differences in trabecular thickness (Tb.Th.). These results are in agreement with previous studies. The stiffness of ROIs from OI bone reached lower values compared to healthy controls and the multilinear fabric-elasticity fits tended to overestimate the stiffness in the lower range. The filtering of highly heterogeneous ROIs removed these low stiffness ROIs and led to similar correlation coefficients for both OI and healthy groups. Finally, the BV/TV and DA matched data revealed no significant differences in fabric-elasticity parameters between OI and healthy individuals. Compared to previous studies, the stiffness constants from the 61 \si{\micro}m resolution HR-pQCT ROIs were lower than for the 36 \si{\micro}m resolution \si{\micro}CT ROIs.

	In conclusion, despite the reduced linear regression parameters found for HR-pQCT images, the fabric-elasticity relationships between OI and healthy individuals are similar when the trabecular bone ROIs are sufficiently homogeneous to perform the mechanical analysis. Since highly heterogenous ROIs coincide with very low BV/TV, we expect them to play a minor role in hFE analysis of distal bone sections.
\end{abstract}

@@

Osteogenesis imperfecta (OI), commonly known as brittle bone disease, is an inherited form of bone fragility \cite{Tournis2018}. OI prevalence is estimated at about  1/13,500 of births, less severe forms not being accounted in this estimation as they are recognized later in life \cite{Lindahl2015}. Regardless, OI is considered as a rare metabolic bone disorder. In most cases, OI is caused by mutations in genes encoding type I collagen (COL1A1 and COL1A2), leading to brittle and fragile bones \cite{LIM2017}, as well as deformed geometry and size in some cases. OI can be categorized according to disease severity \cite{Mortier2019}, where most familiar forms are:
\begin{itemize}
	\item Type I: mild
	\item Type II: perinatally lethal
	\item Type III: most severe surviving form
	\item Type IV: intermediate severity
\end{itemize}

@@

High resolution peripheral quantitative computed tomography (HR-pQCT) reconstructions allow \textit{in vivo} assessment of cortical and trabecular architecture and volumetric bone mineral density (BMD) at the distal radius and distal tibia \cite{Boutroy2005}. Moreover, HR-pQCT reconstructions can be used for finite element analysis (FEA) to predict mechanical properties, such as bone stiffness and strength \cite{Boutroy2008}. Homogenized finite element (hFE) is based on BV/TV and anisotropy information (fabric) from HR-pQCT reconstructions and can be used to assess bone strength within reasonable computation effort \cite{Pahr2009}. High correlations were found between patient-specific hFE and mechanical compression experiments of freshly frozen human samples at the distal radius \cite{Varga2011,Hosseini2017,AriasMoreno2019}. Thus, it could be legitimate to use hFE on OI affected bone, for estimation of bone strength and potentially associated fracture risk. However, HR-pQCT-based hFE relies on fabric-elasticity relationships. Therefore, the present study aims to compare trabecular bone microstructure of healthy and OI bone samples and to test the hypothesis of similar fabric-elasticity relationships.

@@

After ROI cleaning, i.e. deletion of bone material regions unconnected to the main structure, the morphological analysis of ROIs was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffst\"{a}tten, Austria). The morphological parameters analyzed were: BV/TV, structural model index (SMI), trabecular number (Tb.N.), trabecular thickness (Tb.Th.), trabecular separation (Tb.Sp.), and the standard deviation of the trabecular separation (Tb.Sp.SD). Moreover, ROI tissue mineral density (TMD) and fabric was evaluated. The fabric tensor $\mathbf{M}$ was computed using mean intercept length (MIL) method \cite{Moreno2014}. It is a positive-definite second-order tensor computed as shown below.

@@

\subsubsection{Participants Matching}
As the groups do not have similar distributions of age and sex, a matching was performed. The matching algorithm computed the Mahalanobis distance between participants with age and sex as covariates. Then, best correspondences between OI and healthy participants were kept and duplicates were dropped (i.e. no replacement)\cite{Stuart2010}. This matching led to identical mean and median age as well as identical sex distribution.

@@

\begin{table*}[b]
	\centering
	\caption{Summary of the number of ROIs used for the different steps of the study.}
	\label{Table1}
	\begin{tabular}{p{0.1\linewidth}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}}
		\toprule
		Data sets & \multicolumn{2}{c}{Original} & \multicolumn{2}{c}{Age \& sex matched} & \multicolumn{2}{c}{CV filtered} & \multicolumn{2}{c}{BV/TV \& DA matched} \\
		\midrule
		Group & Healthy & OI & Healthy & OI & Healthy & OI & Healthy & OI \\
		Individuals & 120 & 49 & 28 & 28 & 119 & 38 & 57 & 32 \\
		ROIs & 720 & 294 & 168 & 168 & 603 & 117 & 82 & 82 \\
		\midrule
		Methods & \multicolumn{2}{c}{Linear regression} & \multicolumn{2}{c}{Statistics} & \multicolumn{2}{c}{Linear regression} & \multicolumn{2}{c}{Linear regression} \\
		\bottomrule
	\end{tabular}
\end{table*}

@@

As the previous studies \cite{Folkestad2012,Kocijan2015,Rolvien2018} have the same age range as our matched groups, we can compare the morphological parameters. The CT imaging system explains most of the differences between the absolute morphological values of the present study compared to the others. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} have performed their measurements on first generation XCT scanners with a voxel size of 82 \si{\micro}m, while we have used a second generation XCT with a voxel size of 61 \si{\micro}m. The work from \citeauthor{Agarwal2016}\cite{Agarwal2016} investigated differences between the two scanner types. They showed that BV/TV, Tb.Th., and Tb.Sp. are higher in second generation XCT scanners and Tb.N. is lower compared to first generation XCT. These findings are in full agreement with our observations. Another potential bias is introduced by the fact that the present study analyses the median values of six cubic ROIs with 5.3 mm side length. This conditions the Tb.N. and Tb.Sp. as they depend on the ROI size. Moreover, conditions imposed for random ROI selection can lead to further biased values, especially for OI patients, as the ROI must contain at least some portion of trabecular bone. Even with the low sample size (2x28 individuals), the statistical tests have shown significant differences between groups, with the CV presenting the most significant difference. The CV shows that heterogeneity of OI trabecular bone is higher compared to healthy control and is consequently the more discriminant parameter. Finally, the significant differences observed in BV/TV and DA, even with matched age and sex, justifies the choice of a variable matching for fabric-elasticity relationships analysis, because the linear regression must be performed on similar ranges to obtain comparable values. \\

>>>>

\author[2,3]{Mahdi Tabatabaei}

\author[2,3]{Bettina M. Willie}

@@

\address[3]{Departement of Pediatric Surgery, McGill University, Montreal, Canada}

@@

	In the present study, the morphology of distal tibiae from 50 adults with OI were compared to 120 healthy controls using second generation HR-pQCT. Six cubic regions of interest (ROIs) were selected per individual in a common anatomical region. A first matching between OI and healthy control group was performed by selecting similar individuals to obtain identical mean and median age and gender distribution. It allowed to perform a first morphometric analysis and compare the outcome with literature. Then, stiffness tensors of the ROIs were computed using  \si{\micro}FE and multiple linear regressions were performed with the Zysset-Curnier orthotropic fabric-elasticity model. %The linear regressions allowed to compare the two groups using five parameters.

@@

	Compared to healthy controls, we found the OI samples to have significantly lower BV/TV and trabecular number (Tb.N.), significantly higher trabecular separation (Tb.Sp.) and trabecular spacing standard deviation (Tb.Sp.SD), but no differences in trabecular thickness (Tb.Th.). These results are in agreement with previous studies. The stiffness of ROIs from OI bone reached lower values compared to healthy controls and the multilinear fabric-elasticity fits tended to overestimate the stiffness in the lower range. The filtering of highly heterogeneous ROIs removed these low stiffness ROIs and lead to similar correlation coefficients for both OI and healthy groups. Finally, the BV/TV and DA matched data revealed no significant differences in fabric-elasticity parameters between OI and healthy individuals. Compared to previous studies, the stiffness constants from the 61 \si{\micro}m resolution HR-pQCT ROIs were lower than for the 36 \si{\micro}m resolution \si{\micro}CT ROIs.

	In conclusion, despite the reduced linear regression parameters found for HR-pQCT images, the fabric-elasticity relationships between OI and healthy individuals are similar when the trabecular bone ROIs are sufficiently homogeneous to perform the mechanical analysis. Since highly heterogenous ROIs coincide with very low BV/TV, we expect them to play a minor role in hFE analysis of distal bone sections or parts.
\end{abstract}

@@

Osteogenesis imperfecta (OI), commonly known as brittle bone disease, is an inherited form of bone fragility \cite{Tournis2018}. OI prevalence is estimated at about  1/13,500 of births, less severe forms not being accounted in this estimation as they are recognized later in life \cite{Lindahl2015}. Therefore, OI is considered as a rare metabolic bone disorder. In most cases, OI is caused by mutations in genes encoding type I collagen (COL1A1 and COL1A2), leading to brittle and fragile bones \cite{LIM2017}, as well as deformed geometry and size in some cases. OI can be categorized according to disease severity \cite{Mortier2019}, where most familiar forms are:
\begin{itemize}
	\item Type I: mild
	\item Type II: perinatally lethal
	\item Type III: most severe surviving form
	\item Type IV: intermediate severity
\end{itemize}

@@

High resolution peripheral quantitative computed tomography (HR-pQCT) reconstructions allow \textit{in vivo} assessment of cortical and trabecular architecture and volumetric bone mineral density (BMD) at the distal radius and distal tibia \cite{Boutroy2005}. Moreover, HR-pQCT reconstructions can be used for finite element analysis (FEA) to predict mechanical properties, such as bone stiffness and strength \cite{Boutroy2008}. Homogenized finite element (hFE) is based on BV/TV and anisotropy information (fabric) from HR-pQCT reconstructions and can be used to assess bone strength within reasonable computation effort \cite{Pahr2009}. High correlations were found between patient-specific hFE and mechanical compression experiments of freshly frozen human samples at the distal radius \cite{Varga2011,AriasMoreno2019}. Thus, it could be legitimate to use hFE on OI affected bone, for estimation of bone strength and potentially associated fracture risk. However, HR-pQCT-based hFE relies on fabric-elasticity relationships. Therefore, the present study aims to compare trabecular bone microstructure of healthy and OI bone samples and to test the hypothesis of similar fabric-elasticity relationships.

@@

After ROI cleaning, i.e. deletion of bone material regions unconnected to the main structure, the morphological analysis of ROIs was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffst\"{a}tten, Austria). The morphological parameters analyzed were: BV/TV, structural model index (SMI), trabecular number (Tb.N.), trabecular thickness (Tb.Th.), trabecular separation (Tb.Sp.), and the standard deviation of the trabecular spacing (Tb.Sp.SD). Moreover, ROI tissue mineral density (TMD) and fabric was evaluated. The fabric tensor $\mathbf{M}$ was computed using mean intercept length (MIL) method \cite{Moreno2014}. It is a positive-definite second-order tensor computed as shown below.

@@

\subsubsection{Participants Matching}
As the groups do not have similar distributions of age and sex, a matching was performed. The matching algorithm computed the Mahalanobis distance between participants with age and gender as covariates. Then, best correspondences between OI and healthy participants were kept and duplicates were dropped (i.e. no replacement)\cite{Stuart2010}. This matching led to identical mean and median age as well as identical gender distribution.

@@

\begin{table*}[b]
	\centering
	\caption{Summary of the number of ROIs used for the different steps of the study.}
	\label{Table1}
	\begin{tabular}{p{0.1\linewidth}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}*{2}{>{\centering\arraybackslash}p{0.075\linewidth}}}
		\toprule
		Data sets & \multicolumn{2}{c}{Original} & \multicolumn{2}{c}{Age \& gender matched} & \multicolumn{2}{c}{CV filtered} & \multicolumn{2}{c}{BV/TV \& DA matched} \\
		\midrule
		Group & Healthy & OI & Healthy & OI & Healthy & OI & Healthy & OI \\
		Individuals & 120 & 49 & 28 & 28 & 119 & 38 & 57 & 32 \\
		ROIs & 720 & 294 & 168 & 168 & 603 & 117 & 82 & 82 \\
		\midrule
		Methods & \multicolumn{2}{c}{Linear regression} & \multicolumn{2}{c}{Statistics} & \multicolumn{2}{c}{Linear regression} & \multicolumn{2}{c}{Linear regression} \\
		\bottomrule
	\end{tabular}
\end{table*}

@@

As the previous studies \cite{Folkestad2012,Kocijan2015,Rolvien2018} have the same age range as our matched groups, we can compare the morphological parameters. The CT imaging system explains most of the differences between the absolute morphological values of the present study compared to the others. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} have performed their measurements on first generation XCT scanners with a voxel size of 82 \si{\micro}m, while we have used a second generation XCT with a voxel size of 61 \si{\micro}m. The work from \citeauthor{Agarwal2016}\cite{Agarwal2016} investigated differences between the two scanner types. They showed that BV/TV, Tb.Th., and Tb.Sp. are higher in second generation XCT scanners and Tb.N. is lower compared to first generation XCT. These findings are in full agreement with our observations. Another potential bias is introduced by the fact that the present study analyses the median values of six cubic ROIs with 5.3 mm side length. This conditions the Tb.N. and Tb.Sp. as they depend on the ROI size. Moreover, conditions imposed for random ROI selection can lead to further biased values, especially for OI patients, as the ROI must contain at least some portion of trabecular bone. Even with the low sample size (2x28 individuals), the statistical tests have shown significant differences between groups, with the CV presenting the most significant difference. The CV shows that heterogeneity of OI trabecular bone is higher compared to healthy control and is consequently the more discriminant parameter. Finally, the significant differences observed in BV/TV and DA, even with matched age and gender, justifies the choice of a variable matching for fabric-elasticity relationships analysis, because the linear regression must be performed on similar ranges to obtain comparable values. \\
